Predicting Absolute Holocellulose and Klason Lignin Contents for Peat Remains Challenging

### 2022-06-01 analysis/paper/001-reply-main.Rmd
Add missing references in manuscript 2 to manuscript 1 (available as preprint).

1. The models described in @Hodgkins.2018 to predict holocellulose and Klason lignin contents from mid infrared spectra, but recomputed in a Bayesian framework. This is the "original Bayesian model" in [---todo:ref]. These models were used to reproduce the computations in @Hodgkins.2018. Here, these models are called "original" models.

2. A modified model to predict each, holocellulose and Klason lignin contents from mid infrared spectra. These are the models called "best binned spectra" models in [---todo:ref]. A brief description of these models is given in section [Computation of modified models to predict organic matter holocellulose and Klason lignin contents from mid infrared spectra]. These models are used to reanalyze holocellulose and Klason lignin depth profiles for the two latitude classes defined in @Hodgkins.2018, as described in section [Depth profiles].

@@

Computation of the modified models using only the mid infrared spectra (MIRS) is described in detail in [---todo:ref]. In brief, we used the same training data as @Hodgkins.2018 and a Bayesian linear regression model which we computed with Markov Chain Monte Carlo (MCMC) sampling using brms [@Burkner.2017; @Burkner.2018] and rstan [@StanDevelopmentTeam.2020] which are both interfaces to Stan [@StanDevelopmentTeam.2021]. In contrast to the original models of @Hodgkins.2018, we (1) used a beta distribution as likelihood function instead of assuming a Gaussian distribution, because holocellulose and Klason lignin contents cannot be negative or larger than 100 mass-%, we (2) did not use manually selected peaks extracted from the spectra, but all the $z$-transformed absorbance values after baseline correction, normalization, and binning (bin width = 20 cm$^{-1}$), and we (3) did not use simple linear regression, but regularized linear regression by using a regularized horseshoe prior [@Piironen.2017a; @Piironen.2017b] for coefficients for spectral variables. For the intercept and for the scale parameter of the beta distribution, we used weakly informative default priors of rstanarm [@Goodrich.2020] and brms [@Burkner.2017; @Burkner.2018].

@@

1. Since representative training data are missing, both the original and modified models cannot be fully validated and prediction uncertainties are probably larger than computed with these models [---todo:ref]. Since representative test data are missing, these additional uncertainties cannot be estimated. We assume that these uncertainties are larger than those induced by parameter uncertainty from models for the depth profiles.

@@

The modified models using only the MIRS were also used to compute median values and confidence intervals mentioned in the main text in section "Klason lignin content and residual enrichment in (sub)tropical peatlands probably are larger". In these computations, we also considered the predictive uncertainty of the original models (the Bayesian formulation described in ---todo:ref).

@@

To create figure 1<!--\@ref(fig:reply-res-p-sites-latitude)--> in the main text, we used the models for the latitudinal gradient and samples from the posterior distribution to compute the predicted average and 95% confidence interval for a range of latitude values and we added predictions for the average contents for each of the sampled cores. Next, we recomputed the models from @Hodgkins.2018 by (1) predicting average peat sample contents with the original models (the Bayesian version described in ---todo:ref), (2) averaging the near surface peat sample average contents for each core, and (3) fitting an ordinary least squares regression line to these average core values. For this linear model, we plotted the confidence interval across latitude and we added the computed average core near-surface contents. Thus, figure 1<!---\@ref(fig:reply-res-p-sites-latitude)---> in the main text compares the graphed values and model prediction from figure 3 in @Hodgkins.2018 with predictions from our modified analysis considering prediction uncertainty from the MIRS-based model part, and the averaging per core.  
Additional sources for differences in predictions and prediction uncertainties in comparison to the analysis in @Hodgkins.2018 are the use of a beta distribution which constraints uncertainty intervals for extreme (near 0 and 100 mass-%) values, and the use of all spectral variables which resulted in reduced prediction uncertainties [---todo:ref].

>>>>

1. The models described in @Hodgkins.2018 to predict holocellulose and Klason lignin contents from mid infrared spectra, but recomputed in a Bayesian framework. This is the "original Bayesian model" in @Teickner.2022a. These models were used to reproduce the computations in @Hodgkins.2018. Here, these models are called "original" models.

2. A modified model to predict each, holocellulose and Klason lignin contents from mid infrared spectra. These are the models called "best binned spectra" models in @Teickner.2022a. A brief description of these models is given in section [Computation of modified models to predict organic matter holocellulose and Klason lignin contents from mid infrared spectra]. These models are used to reanalyze holocellulose and Klason lignin depth profiles for the two latitude classes defined in @Hodgkins.2018, as described in section [Depth profiles].

@@

Computation of the modified models using only the mid infrared spectra (MIRS) is described in detail in @Teickner.2022a. In brief, we used the same training data as @Hodgkins.2018 and a Bayesian linear regression model which we computed with Markov Chain Monte Carlo (MCMC) sampling using brms [@Burkner.2017; @Burkner.2018] and rstan [@StanDevelopmentTeam.2020] which are both interfaces to Stan [@StanDevelopmentTeam.2021]. In contrast to the original models of @Hodgkins.2018, we (1) used a beta distribution as likelihood function instead of assuming a Gaussian distribution, because holocellulose and Klason lignin contents cannot be negative or larger than 100 mass-%, we (2) did not use manually selected peaks extracted from the spectra, but all the $z$-transformed absorbance values after baseline correction, normalization, and binning (bin width = 20 cm$^{-1}$), and we (3) did not use simple linear regression, but regularized linear regression by using a regularized horseshoe prior [@Piironen.2017a; @Piironen.2017b] for coefficients for spectral variables. For the intercept and for the scale parameter of the beta distribution, we used weakly informative default priors of rstanarm [@Goodrich.2020] and brms [@Burkner.2017; @Burkner.2018].

@@

1. Since representative training data are missing, both the original and modified models cannot be fully validated and prediction uncertainties are probably larger than computed with these models [@Teickner.2022a]. Since representative test data are missing, these additional uncertainties cannot be estimated. We assume that these uncertainties are larger than those induced by parameter uncertainty from models for the depth profiles.

@@

The modified models using only the MIRS were also used to compute median values and confidence intervals mentioned in the main text in section "Klason lignin content and residual enrichment in (sub)tropical peatlands probably are larger". In these computations, we also considered the predictive uncertainty of the original models (the Bayesian formulation described in @Teickner.2022a).

@@

To create figure 1<!--\@ref(fig:reply-res-p-sites-latitude)--> in the main text, we used the models for the latitudinal gradient and samples from the posterior distribution to compute the predicted average and 95% confidence interval for a range of latitude values and we added predictions for the average contents for each of the sampled cores. Next, we recomputed the models from @Hodgkins.2018 by (1) predicting average peat sample contents with the original models (the Bayesian version described in @Teickner.2022a), (2) averaging the near surface peat sample average contents for each core, and (3) fitting an ordinary least squares regression line to these average core values. For this linear model, we plotted the confidence interval across latitude and we added the computed average core near-surface contents. Thus, figure 1<!---\@ref(fig:reply-res-p-sites-latitude)---> in the main text compares the graphed values and model prediction from figure 3 in @Hodgkins.2018 with predictions from our modified analysis considering prediction uncertainty from the MIRS-based model part, and the averaging per core.  
Additional sources for differences in predictions and prediction uncertainties in comparison to the analysis in @Hodgkins.2018 are the use of a beta distribution which constraints uncertainty intervals for extreme (near 0 and 100 mass-%) values, and the use of all spectral variables which resulted in reduced prediction uncertainties [@Teickner.2022a].

### 2022-06-01 analysis/paper/002-reply-main.Rmd
Add DOI for zenodo upload.
Add funding information.
Add missing references in manuscript 2 to manuscript 1 (available as preprint).

In a recent evaluation of the models to predict peat holocellulose and Klason lignin contents presented in @Hodgkins.2018, we found that the training data used for the models (leaves, paper products) are not representative for peat [---todo: ref]. The peat samples are more decomposed than the training data samples. Decomposition typically causes distinct features in mid infrared spectra which have complex effects on the applicability of spectral prediction models [---todo:ref]. Since these features are not present in the training data, predictions of the original model are extrapolations. Following our re-evaluation, this causes large uncertainties in model predictions for both holocellulose and Klason lignin, and a comparatively large bias towards smaller Klason lignin contents, especially for more decomposed peat with Klason lignin contents >~30 mass-% [---todo:ref]. Because of these uncertainties and biases, the analyses of @Hodgkins.2018 cannot be interpreted in terms of absolute holocellulose and Klason lignin contents.  
Instead, we suggest to interpret differences in peat chemistry presented by @Hodgkins.2018 not in terms of specific organic matter fractions (i.e. holocellulose and Klason lignin), but rather in terms of more qualitative concepts typically used in studies interpreting mid infrared spectra of peat (e.g. @Artz.2008; @Broder.2012; @Tfaily.2014; @Biester.2014), i.e. as "carbohydrates" and "aromatics". We assume that under such an interpretation, the conclusions of @Hodgkins.2018 probably are valid.  
The proposed reinterpretation is possible if values given in @Hodgkins.2018 are only interpreted in terms of the normalized peak heights used in the original prediction models, and not as absolute holocellulose and Klason lignin contents. Each model is a linear model with one independent variable (the normalized height of one peak); therefore a direct reinterpretation of the predicted absolute contents in terms of normalized peak heights is possible if the figure axes are relabeled as "Relative content of carbohydrates/aromatics" instead of "Estimated % carbohydrates/aromatics" (implying an absolute content of holocellulose and Klason lignin, respectively) [@Hodgkins.2018].   
An important open question to this approach of interpreting large-scale gradients in peat chemistry based on mid infrared spectra should be addressed in future studies: Are differences in specific mid infrared spectra peaks reliable indicators for differences in a specific concept for peat chemistry? For example, are differences in the peak around 1040 cm$^{-1}$ indicative for differences in carbohydrate contents (also across larger scales)?  
A similar problem arose in the discussion on the use of the C/N ratio as a good indicator for peat mass loss. This may only hold true if there have been no or negligible shifts in the vegetation composition which causes differences in the C/N ratio already in initial plant litter (e.g. @Biester.2014). For mid infrared spectra, such knowledge still has to be acquired in future studies, especially if large environmental gradients are considered as in @Hodgkins.2018.

@@

We suggest that @Hodgkins.2018 have underestimated peat Klason lignin contents, especially in (sub)tropical peat. As described above, the original model for Klason lignin underestimates peat Klason lignin contents. A modified model has less bias [---todo:ref]. In spite of several remaining limitations (see below and the supplementary information), we argue that this modified model can serve as rough indicator for the underestimation of peat Klason lignin contents by the original model. We therefore present values to provide tentative estimates for the underestimation of Klason lignin contents, but we do not claim that these estimates are optimal.  
With this modified model, near-surface ($\le50$ cm) peat Klason lignin contents are on average ~10 to 15 mass-% larger across the latitudinal gradient (figure \@ref(fig:reply-res-p-sites-latitude)). Specifically, average (sub)tropical near-surface peat ($<45 ^\circ$N, $\le50$ cm) Klason lignin contents are `r d_peat_pred_best_diff_mean %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude & surface_peat) %>% dplyr::pull(y_hat_diff) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude & surface_peat) %>% dplyr::pull(y_hat_diff) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-% (median, lower and upper 95% prediction interval limit) larger than with the original model. There is also tentative indication for on average larger differences between deep (sub)tropical ($<45 ^\circ$N) and high-latitude ($\ge45 ^\circ$N) peat (figure \@ref(fig:reply-res-p-depth-profiles-latitude)).    
We also suggest that the residual enrichment of Klason lignin during decomposition probably has been underestimated. A rough estimate for the residual enrichment of Klason lignin during decomposition is the difference in Klason lignin content between near-surface peat ($\le50$ cm) and deeper peat. With the modified model, this difference is on average for (sub)tropical peatlands `r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude) %>% dplyr::pull(res) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude) %>% dplyr::pull(res) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-% larger (figure \@ref(fig:reply-res-p-depth-profiles-latitude)). For high latitude peatlands the difference is smaller on average (`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & high_latitude) %>% dplyr::pull(res) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & high_latitude) %>% dplyr::pull(res) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-%).  
Consequently, in general --- and especially in (sub)tropical peatlands --- the average residual enrichment of Klason lignin due to decomposition probably has been underestimated by the original model. A consequence of this in light of climate change is that high latitude peat deposits may be more prone to shifts in decomposition rates than suggested by the original results.

@@

Besides a reinterpretation in terms of qualitative concepts for organic matter recalcitrance, we also believe that a more comprehensive quantification and propagation of uncertainties is important to further scrutinize the results of @Hodgkins.2018. In the original analysis, the following uncertainties are omitted: (1) As described above, the training dataset is not representative for peat and therefore predictions are biased. This uncertainty is not considered in the analyses of @Hodgkins.2018. (2) The second source of uncertainty which was omitted is the prediction uncertainty of the models used to predict holocellulose and Klason lignin contents. (3) A third source of uncertainty not considered yet is that in computing the regression model to analyze the latitudinal gradient based on average core near-surface contents, the uncertainty for these average values was not propagated.  
The first two sources of uncertainty can be omitted if only normalized peak heights (as qualitative indicators) are used instead of predicted holocellulose and Klason lignin contents, as described above. But the third source of uncertainty should also be considered when reinterpreting the results in terms of area normalized peak heights.  
To support our concerns, we reanalyzed the latitudinal gradient for holocellulose and Klason lignin, respectively, using the modified models described in [---todo:ref] and fully considering the last two sources of uncertainty using a Bayesian approach (described in more detail in the supplementary information). With the modified models, the estimated slopes for the latitudinal gradient are more uncertain (95% confidence intervals are [`r s1_res_holocellulose %>% dplyr::pull(b_latitude) %>% quantile(probs = c(0.025, 0.975)) %>% round(1) %>% paste0(collapse = ",")`] and [`r s1_res_klason_lignin %>% dplyr::pull(b_latitude) %>% quantile(probs = c(0.025, 0.975)) %>% round(1) %>% paste0(collapse = ",")`] for holocellulose and Klason lignin, respectively; figure \@ref(fig:reply-res-p-sites-latitude) and supporting figure 3<!---\@ref(fig:reply-res-p-latitude-poc-slope)--->). A similar reanalysis of the depth profiles between the two latitude categories shows that differences are also more uncertain (figure \@ref(fig:reply-res-p-depth-profiles-latitude)). Our analysis therefore indicates larger uncertainties both in the depth profiles and the latitudinal gradient. Considering these uncertainties is important when peat properties predicted with a model are analyzed.
Again, the values presented here are only tentative because the modified models probably are also biased.

@@

Notwithstanding our critical remarks, the idea of @Hodgkins.2018 to relate mid infrared spectra to precisely quantifiable and defined organic matter fractions is important. Such fractions are relatively well understood and therefore allow us to better deduce what effects climate shifts can have on peat decomposition and recalcitrance. It is much more informative if we are able to precisely estimate a change in Klason lignin contents of e.g. 5 mass-% if average annual temperatures increase by 1$^{\circ}$C than if we are only able to make qualitative statements that the relative amount of aromatics (without describing in detail what this fraction means) increases, as is currently the case.  
When interpreting mid infrared spectra, distinguishing between precisely quantifiable concepts for organic matter chemistry, such as holocellulose and Klason lignin contents, and qualitative concepts which currently are used in interpreting peat MIRS, such as "carbohydrates and aromatics" (without specifying how this fraction could actually be measured), is crucial: Models relying on quantitative concepts are more testable and allow to actually quantify fractional changes in organic matter chemistry, whereas models built on qualitative concepts can only be refuted if there are blatant discrepancies relative to expectations (i.e. a decrease in "carbohydrates" with latitude). Ultimately, we need more precise models for peat variations in chemistry than can be developed based on qualitative concepts.  
We suggest the following steps to gain deeper insights into latitudinal gradient in peat chemistry: As outlined above, we suggest to interpret the results of @Hodgkins.2018 in terms of qualitative concepts for peat chemistry, i.e. "carbohydrates" and "aromatics". These suggestions also apply to a follow-up study using the same models [@Verbeke.2022]. Further studies are needed to check if specific peaks are good indicators for these qualitative concepts for peat chemistry, how the relation may be confounded e.g. by vegetation composition or other organic matter fractions, as could be the case especially for peaks related to aromatics [---todo:ref]. This could be done by computing models to reliably predict holocellulose and Klason lignin contents of peat from mid infrared spectra and analyzing which spectral variables are relevant for this purpose and if there are confounding variables. Such models will also help us to get more precise insights into the degree to which high latitude peat accumulation may be decreased by decomposition due to rising temperatures, because they allow replacing more costly measurements. Finally, future studies should carefully consider uncertainties and fully propagate these into model estimates. Predicting absolute peat holocellulose and Klason lignin contents from infrared spectra remains challenging.

@@

We like to thank Jeffrey Chanton and Suzanne Hodgkins for a constructive dicussion which greatly helped to improve a first version of this manuscript.

@@

>>>>

In a recent evaluation of the models to predict peat holocellulose and Klason lignin contents presented in @Hodgkins.2018, we found that the training data used for the models (leaves, paper products) are not representative for peat [@Teickner.2022a]. The peat samples are more decomposed than the training data samples. Decomposition typically causes distinct features in mid infrared spectra which have complex effects on the applicability of spectral prediction models [@Teickner.2022a]. Since these features are not present in the training data, predictions of the original model are extrapolations. Following our re-evaluation, this causes large uncertainties in model predictions for both holocellulose and Klason lignin, and a comparatively large bias towards smaller Klason lignin contents, especially for more decomposed peat with Klason lignin contents >~30 mass-% [@Teickner.2022a]. Because of these uncertainties and biases, the analyses of @Hodgkins.2018 cannot be interpreted in terms of absolute holocellulose and Klason lignin contents.  
Instead, we suggest to interpret differences in peat chemistry presented by @Hodgkins.2018 not in terms of specific organic matter fractions (i.e. holocellulose and Klason lignin), but rather in terms of more qualitative concepts typically used in studies interpreting mid infrared spectra of peat (e.g. @Artz.2008; @Broder.2012; @Tfaily.2014; @Biester.2014), i.e. as "carbohydrates" and "aromatics". We assume that under such an interpretation, the conclusions of @Hodgkins.2018 probably are valid.  
The proposed reinterpretation is possible if values given in @Hodgkins.2018 are only interpreted in terms of the normalized peak heights used in the original prediction models, and not as absolute holocellulose and Klason lignin contents. Each model is a linear model with one independent variable (the normalized height of one peak); therefore a direct reinterpretation of the predicted absolute contents in terms of normalized peak heights is possible if the figure axes are relabeled as "Relative content of carbohydrates/aromatics" instead of "Estimated % carbohydrates/aromatics" (implying an absolute content of holocellulose and Klason lignin, respectively) [@Hodgkins.2018].   
An important open question to this approach of interpreting large-scale gradients in peat chemistry based on mid infrared spectra should be addressed in future studies: Are differences in specific mid infrared spectra peaks reliable indicators for differences in a specific concept for peat chemistry? For example, are differences in the peak around 1040 cm$^{-1}$ indicative for differences in carbohydrate contents (also across larger scales)?  
A similar problem arose in the discussion on the use of the C/N ratio as a good indicator for peat mass loss. This may only hold true if there have been no or negligible shifts in the vegetation composition which causes differences in the C/N ratio already in initial plant litter (e.g. @Biester.2014). For mid infrared spectra, such knowledge still has to be acquired in future studies, especially if large environmental gradients are considered as in @Hodgkins.2018.

@@

We suggest that @Hodgkins.2018 have underestimated peat Klason lignin contents, especially in (sub)tropical peat. As described above, the original model for Klason lignin underestimates peat Klason lignin contents. A modified model has less bias [@Teickner.2022a]. In spite of several remaining limitations (see below and the supplementary information), we argue that this modified model can serve as rough indicator for the underestimation of peat Klason lignin contents by the original model. We therefore present values to provide tentative estimates for the underestimation of Klason lignin contents, but we do not claim that these estimates are optimal.  
With this modified model, near-surface ($\le50$ cm) peat Klason lignin contents are on average ~10 to 15 mass-% larger across the latitudinal gradient (figure \@ref(fig:reply-res-p-sites-latitude)). Specifically, average (sub)tropical near-surface peat ($<45 ^\circ$N, $\le50$ cm) Klason lignin contents are `r d_peat_pred_best_diff_mean %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude & surface_peat) %>% dplyr::pull(y_hat_diff) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude & surface_peat) %>% dplyr::pull(y_hat_diff) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-% (median, lower and upper 95% prediction interval limit) larger than with the original model. There is also tentative indication for on average larger differences between deep (sub)tropical ($<45 ^\circ$N) and high-latitude ($\ge45 ^\circ$N) peat (figure \@ref(fig:reply-res-p-depth-profiles-latitude)).    
We also suggest that the residual enrichment of Klason lignin during decomposition probably has been underestimated. A rough estimate for the residual enrichment of Klason lignin during decomposition is the difference in Klason lignin content between near-surface peat ($\le50$ cm) and deeper peat. With the modified model, this difference is on average for (sub)tropical peatlands `r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude) %>% dplyr::pull(res) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & !high_latitude) %>% dplyr::pull(res) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-% larger (figure \@ref(fig:reply-res-p-depth-profiles-latitude)). For high latitude peatlands the difference is smaller on average (`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & high_latitude) %>% dplyr::pull(res) %>% median() %>% magrittr::multiply_by(100) %>% round(0)` [`r d_peat_pred_best_diff_mean1 %>% dplyr::mutate(res = y_hat_diff - y_hat_or_diff) %>% dplyr::filter(variable_y == "Klason lignin" & high_latitude) %>% dplyr::pull(res) %>% quantile(probs = c(0.025, 0.975)) %>% magrittr::multiply_by(100) %>% round(0) %>% paste(collapse = ",")`] mass-%).  
Consequently, in general --- and especially in (sub)tropical peatlands --- the average residual enrichment of Klason lignin due to decomposition probably has been underestimated by the original model. A consequence of this in light of climate change is that high latitude peat deposits may be more prone to shifts in decomposition rates than suggested by the original results.

@@

Besides a reinterpretation in terms of qualitative concepts for organic matter recalcitrance, we also believe that a more comprehensive quantification and propagation of uncertainties is important to further scrutinize the results of @Hodgkins.2018. In the original analysis, the following uncertainties are omitted: (1) As described above, the training dataset is not representative for peat and therefore predictions are biased. This uncertainty is not considered in the analyses of @Hodgkins.2018. (2) The second source of uncertainty which was omitted is the prediction uncertainty of the models used to predict holocellulose and Klason lignin contents. (3) A third source of uncertainty not considered yet is that in computing the regression model to analyze the latitudinal gradient based on average core near-surface contents, the uncertainty for these average values was not propagated.  
The first two sources of uncertainty can be omitted if only normalized peak heights (as qualitative indicators) are used instead of predicted holocellulose and Klason lignin contents, as described above. But the third source of uncertainty should also be considered when reinterpreting the results in terms of area normalized peak heights.  
To support our concerns, we reanalyzed the latitudinal gradient for holocellulose and Klason lignin, respectively, using the modified models described in @Teickner.2022a and fully considering the last two sources of uncertainty using a Bayesian approach (described in more detail in the supplementary information). With the modified models, the estimated slopes for the latitudinal gradient are more uncertain (95% confidence intervals are [`r s1_res_holocellulose %>% dplyr::pull(b_latitude) %>% quantile(probs = c(0.025, 0.975)) %>% round(1) %>% paste0(collapse = ",")`] and [`r s1_res_klason_lignin %>% dplyr::pull(b_latitude) %>% quantile(probs = c(0.025, 0.975)) %>% round(1) %>% paste0(collapse = ",")`] for holocellulose and Klason lignin, respectively; figure \@ref(fig:reply-res-p-sites-latitude) and supporting figure 3<!---\@ref(fig:reply-res-p-latitude-poc-slope)--->). A similar reanalysis of the depth profiles between the two latitude categories shows that differences are also more uncertain (figure \@ref(fig:reply-res-p-depth-profiles-latitude)). Our analysis therefore indicates larger uncertainties both in the depth profiles and the latitudinal gradient. Considering these uncertainties is important when peat properties predicted with a model are analyzed.
Again, the values presented here are only tentative because the modified models probably are also biased.

@@

Notwithstanding our critical remarks, the idea of @Hodgkins.2018 to relate mid infrared spectra to precisely quantifiable and defined organic matter fractions is important. Such fractions are relatively well understood and therefore allow us to better deduce what effects climate shifts can have on peat decomposition and recalcitrance. It is much more informative if we are able to precisely estimate a change in Klason lignin contents of e.g. 5 mass-% if average annual temperatures increase by 1$^{\circ}$C than if we are only able to make qualitative statements that the relative amount of aromatics (without describing in detail what this fraction means) increases, as is currently the case.  
When interpreting mid infrared spectra, distinguishing between precisely quantifiable concepts for organic matter chemistry, such as holocellulose and Klason lignin contents, and qualitative concepts which currently are used in interpreting peat MIRS, such as "carbohydrates and aromatics" (without specifying how this fraction could actually be measured), is crucial: Models relying on quantitative concepts are more testable and allow to actually quantify fractional changes in organic matter chemistry, whereas models built on qualitative concepts can only be refuted if there are blatant discrepancies relative to expectations (i.e. a decrease in "carbohydrates" with latitude). Ultimately, we need more precise models for peat variations in chemistry than can be developed based on qualitative concepts.  
We suggest the following steps to gain deeper insights into latitudinal gradient in peat chemistry: As outlined above, we suggest to interpret the results of @Hodgkins.2018 in terms of qualitative concepts for peat chemistry, i.e. "carbohydrates" and "aromatics". These suggestions also apply to a follow-up study using the same models [@Verbeke.2022]. Further studies are needed to check if specific peaks are good indicators for these qualitative concepts for peat chemistry, how the relation may be confounded e.g. by vegetation composition or other organic matter fractions, as could be the case especially for peaks related to aromatics [@Teickner.2022a]. This could be done by computing models to reliably predict holocellulose and Klason lignin contents of peat from mid infrared spectra and analyzing which spectral variables are relevant for this purpose and if there are confounding variables. Such models will also help us to get more precise insights into the degree to which high latitude peat accumulation may be decreased by decomposition due to rising temperatures, because they allow replacing more costly measurements. Finally, future studies should carefully consider uncertainties and fully propagate these into model estimates. Predicting absolute peat holocellulose and Klason lignin contents from infrared spectra remains challenging.

@@

We like to thank Jeffrey Chanton and Suzanne Hodgkins for a constructive dicussion which greatly helped to improve a first version of this manuscript. This study was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) grant no. KN 929/23-1 to Klaus-Holger Knorr and grant no. PE 1632/18-1 to Edzer Pebesma.

@@

# Data and Code availability {.unnumbered}

Data and code to reproduce our analyses are available via <https://doi.org/10.5281/zenodo.6325760.>
